Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} Packages

### 2016-09-20 jss/Deveau.tex
very very minor changes.

%% for pretty printing and a nice hypersummary also set:
\author{by Paul Deveau, Emmanuel Barillot, Valentina Boeva, Andrei Zinovyev and Eric Bonnet} %% comma-separated
\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} packages} %% without formatting
%\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)
\maketitle
%% publication information

@@

In the simple example detailed below, we create a NaviCell session, then load
an expression data set from a local (tab-delimited) file. The data represent
gene expression measured in a prostate cancer cell line resistant to hormonal
treatment (agressive), and is taken from the Cell Line Encyclopedia project
\citep{barretina2012cancer}. We visualize the data values on the Cell Cycle map
(the default map), using heat maps. With this visualization mode, gene
expression values are represented as a color gradient (green to red) in squares
positioned next to the entities where the gene has been mapped
(Figure~\ref{fig:du145}). Note that the map is displayed in a browser and is
\emph{interactive}, i.e users can zoom in to display more information and for
example look in what reactions are involved the genes selected to be displayed,
and lots of other informations (see \cite{bonnet2015navicell} and
\cite{kuperstein2015atlas} for more details).

@@

Malignant gliomas, the most common subtype of primary brain tumors, are
aggressive, highly invasive, and neurologically destructive tumors considered
to be among the deadliest of human cancers. In its most aggressive
manifestation, glioblastoma (GBM), median survival ranges from 9 to 12 months,
despite maximum treatment efforts \citep{maher2001malignant}. In this study we
have analyzed whole-genome mutation data generated by the TCGA project on
hundreds of patients. More specifically, we parsed the MAF (Mutation Annotation
Format) GBM files produced by different sequencing centers to count and
calculate gene mutation frequencies. We kept the mutations having a status
likely to disturb the target protein's function (i.e, Frame\_Shift\_Del,
Nonstop\_Mutation, In\_Frame\_Del, In\_Frame\_Ins, Missense\_Mutation,
Nonsense\_Mutation, Splice\_Site, Translation\_Start\_Site). In total, we
collected mutations for more than 13,000 genes in a total of 379 mutated
samples. In order to retain the most frequently mutated genes, we calculated
frequencies across all mutated samples, and kept genes having a frequency
greater than 1\% (3,293 genes). We further labelled genes having a frequency
greater than 1\% and less than 5\% as "1" and genes highly mutated (frequency
higher than 5\%) as "2".

>>>>

%% for pretty printing and a nice hypersummary also set:
\author{by Paul Deveau, Emmanuel Barillot, Valentina Boeva, Andrei Zinovyev and Eric Bonnet} %% comma-separated
\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} Packages} %% without formatting
%\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)
\maketitle
%% publication information

@@

In the simple example detailed below, we create a NaviCell session, then load
an expression data set from a local (tab-delimited) file. The data represent
gene expression measured in a prostate cancer cell line resistant to hormonal
treatment (agressive), and is taken from the Cell Line Encyclopedia project
\citep{barretina2012cancer}. We visualize the data values on the Cell Cycle map
(the default map), using heat maps. With this visualization mode, gene
expression values are represented as a color gradient (green to red) in squares
positioned next to the entities where the gene has been mapped
(Figure~\ref{fig:du145}). Note that the map is displayed in a browser and is
\emph{interactive}, i.e.\ users can zoom in to display more information and for
example look in what reactions are involved the genes selected to be displayed,
and lots of other informations (see \cite{bonnet2015navicell} and
\cite{kuperstein2015atlas} for more details).

@@

Malignant gliomas, the most common subtype of primary brain tumors, are
aggressive, highly invasive, and neurologically destructive tumors considered
to be among the deadliest of human cancers. In its most aggressive
manifestation, glioblastoma (GBM), median survival ranges from 9 to 12 months,
despite maximum treatment efforts \citep{maher2001malignant}. In this study we
have analyzed whole-genome mutation data generated by the TCGA project on
hundreds of patients. More specifically, we parsed the MAF (Mutation Annotation
Format) GBM files produced by different sequencing centers to count and
calculate gene mutation frequencies. We kept the mutations having a status
likely to disturb the target protein's function (i.e.\, Frame\_Shift\_Del,
Nonstop\_Mutation, In\_Frame\_Del, In\_Frame\_Ins, Missense\_Mutation,
Nonsense\_Mutation, Splice\_Site, Translation\_Start\_Site). In total, we
collected mutations for more than 13,000 genes in a total of 379 mutated
samples. In order to retain the most frequently mutated genes, we calculated
frequencies across all mutated samples, and kept genes having a frequency
greater than 1\% (3,293 genes). We further labelled genes having a frequency
greater than 1\% and less than 5\% as "1" and genes highly mutated (frequency
higher than 5\%) as "2".
